Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-10-07 formally.tex
updated implementation section for AE
prepping paper for artifact evaluation

Our overview below omits certain ``uninteresting'' details. The supplement
includes the complete collection of rules, in definitional order. These
rules, along with the proofs of all of the metatheorems listed, have been
mechanized in Agda \cite{norell:thesis}, also in the supplement. All the
judgements that mention contexts described below enjoy the standard
structural properties of weakening, contraction, and exchange, and this too
is proven in the mechanization. We will summarize this effort in
Sec. \ref{sec:mech}.

@@

Types and expressions that contain no holes are \emph{complete types} and
\emph{complete expressions}, respectively. Formally, we can derive
$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
$\hexp$ is complete (see supplement.) We are not concerned here with
defining a dynamic semantics for Hazelnut, but the dynamics for complete
H-expressions would be entirely standard (in Sec. \ref{sec:rw} we discuss
the multi-dimensional design space around the question of evaluating
incomplete H-expressions.) For completeness, a mapping from the
bidirectional type system to a declarative one, and a proof of type safety
for complete terms in that system, is included in the mechanization.

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}
\begin{figure}
\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
\begin{subequations}\label{eqns:consistency}
\begin{mathpar}

\inferrule{ }{
  \tcompat{\tehole}{\htau}
}
~~~~~~~~
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}
~~~~~~~~
\inferrule{ }{
  \tcompat{\htau}{\htau}
}
~~~~~~~~
\inferrule{
  \tcompat{\htau_1}{\htau_1'}\\
  \tcompat{\htau_2}{\htau_2'}
}{
  \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
}~~~~~~~~\text{(\ref*{eqns:consistency}a-d)}
\end{mathpar}
\end{subequations}
\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{subequations}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}
\end{equation}
\end{minipage}
\end{subequations}
\caption{H-type consistency and matched arrow types.}
\label{fig:type-consistency}
\end{figure}

@@

Hazelnut's static semantics is organized as a \emph{bidirectional type
  system}
\cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial}
around the two mutually defined judgements in Figure
\ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in
\domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
identify the context up to exchange and adopt the standard identification
convention for formal structures that differ only up to alpha-equivalence.

@@

The action semantics includes a subsumption rule similar to the subsumption
rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type,
then we defer to the synthetic action judgement, as long as the type of the
resulting cursor erasure is consistent with the type provided for analysis.
The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem
\ref{thrm:actsafe} goes through by induction and static subsumption,
i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be
the rule of last resort.

@@

\subsection{Determinism}
A last useful property to consider is \emph{action determinism}, i.e. that
performing an action produces a unique result. The following theorem establishes determinism for type actions:
\begin{theorem}[Type Action Determinism]
\label{thrm:type-actdet} If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\end{theorem}

@@

The mechanization, discussed next, defines the predicates $\subminsyn{\mathcal{D}}$ and $\subminana{\mathcal{D}}$. In addition, it defines a mapping from
any derivation into a corresponding subsumption-minimal derivation. Implementors of Hazelnut are expected only to implement this subsumption-minimal fragment.

>>>>

Our overview below omits certain ``uninteresting'' details. The appendix
includes the complete collection of rules, in definitional order. These
rules, along with the proofs of all of the metatheorems discussed, have been
mechanized in Agda \cite{norell:thesis} (see Sec. \ref{sec:mech}.)

@@

Types and expressions that contain no holes are \emph{complete types} and
\emph{complete expressions}, respectively. Formally, we can derive
$\hcomplete{\htau}$ when $\htau$ is complete, and $\hcomplete{\hexp}$ when
$\hexp$ is complete (see supplement.) We are not concerned here with
defining a dynamic semantics for Hazelnut, but the dynamics for complete
H-expressions would be entirely standard because complete expressions correspond  immediately to those of the typed lambda calculus with numbers (in Sec. \ref{sec:rw} we discuss
the multi-dimensional design space around the question of evaluating
incomplete H-expressions.)

\begin{figure}[t]
$\arraycolsep=4pt\begin{array}{lllllll}
\mathsf{HTyp} & \htau & ::= &
  \tarr{\htau}{\htau} ~\vert~
  \tnum ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  \hexp : \htau ~\vert~
  x ~\vert~
  \hlam{x}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  \hnum{n} ~\vert~
  \hadd{\hexp}{\hexp} ~\vert~
  \hehole ~\vert~
  \hhole{\hexp}
\end{array}$
\caption{Syntax of H-types and H-expressions. Metavariable $x$ ranges over variables and $n$ ranges over numerals.}
\label{fig:hexp-syntax}
\end{figure}
\begin{figure}
\noindent\fbox{$\tcompat{\htau}{\htau'}$}~~\text{$\tau$ and $\tau'$ are consistent}
\begin{subequations}\label{eqns:consistency}
\begin{mathpar}
\inferrule{ }{
  \tcompat{\tehole}{\htau}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\tehole}
}
~~~~~
\inferrule{ }{
  \tcompat{\htau}{\htau}
}
~~~~~
\inferrule{
  \tcompat{\htau_1}{\htau_1'}\\
  \tcompat{\htau_2}{\htau_2'}
}{
  \tcompat{(\tarr{\htau_1}{\htau_2})}{(\tarr{\htau_1'}{\htau_2'})}
}~~~~\text{\hspace{-2px}(\ref*{eqns:consistency}a-d)}
\end{mathpar}
\end{subequations}
\fbox{$\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$}~~\text{$\tau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{subequations}
\begin{minipage}{0.5\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{equation}
\end{minipage}
\begin{minipage}{0.48\linewidth}
\begin{equation}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}
\end{equation}
\end{minipage}
\end{subequations}
\caption{H-type consistency and matched arrow types.}
\label{fig:type-consistency}
\end{figure}

@@

Hazelnut's static semantics is organized as a \emph{bidirectional type
  system}
\cite{Pierce:2000:LTI:345099.345100,DBLP:conf/icfp/DaviesP00,DBLP:conf/tldi/ChlipalaPH05,bidi-tutorial}
around the two mutually defined judgements in Figure
\ref{fig:ana-syn}. Typing contexts, $\hGamma$, map each variable $x \in
\domof{\hGamma}$ to a hypothesis $x : \htau$ in the usual manner. We
identify the context up to exchange and contraction and adopt the standard identification
convention for formal structures that differ only up to alpha-renaming of bound variables. All judgements that mention a context obey weakening.

@@

The action semantics includes a subsumption rule similar to the subsumption
rule, Rule (\ref{rule:ana-subsume}), in the statics:
\begin{equation}\label{rule:action-subsume}
  \inferrule{
    \hsyn{\hGamma}{\removeSel{\zexp}}{\htau'}\\
    \performSyn{\hGamma}{\zexp}{\htau'}{\alpha}{\zexp'}{\htau''}\\
    \tcompat{\htau}{\htau''}
  }{
    \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}
  }
\end{equation}
In other words, if the cursor erasure of the edit state synthesizes a type,
then we defer to the synthetic action judgement, as long as the type of the
resulting cursor erasure is consistent with the type provided for analysis.
The case involving Rule (\ref{rule:action-subsume}) in the proof of Theorem
\ref{thrm:actsafe} goes through by induction and static subsumption,
i.e. Rule (\ref{rule:ana-subsume}). Algorithmically, subsumption should be
the rule of last resort (see Sec. \ref{sec:determinism}.)

@@

\subsection{Determinism}\label{sec:determinism}
A last useful property to consider is \emph{action determinism}, i.e. that
performing an action produces a unique result. The following theorem establishes determinism for type actions:
\begin{theorem}[Type Action Determinism]
\label{thrm:type-actdet} If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\end{theorem}

@@

The mechanization, discussed next, defines the predicates $\subminsyn{\mathcal{D}}$ and $\subminana{\mathcal{D}}$. In addition, it defines a mapping from
any derivation into a corresponding subsumption-minimal derivation. Implementations of Hazelnut need only implement this subsumption-minimal fragment.
